---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
The data provided was inside the repository in a file called "activity.zip"

The variables in the dataset are the following:

* **steps**: Number of steps taking in a 5-minute interval (missing values are coded as NA)
* **date**: The date on which the measurement was taken in YYYY-MM-DD format
* **interval**: Identifier for the 5-minute interval in which measurement was taken

the dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.

- Unzip the data file

``` {r}
unzip(zipfile = "activity.zip", files = "activity.csv", overwrite = TRUE)
```
- Load and store the datafile in a dataframe object called "activitydata"

```{r}
activitydata <-read.table(file = "activity.csv", header = TRUE, sep = ",", na.strings = "NA")

str(activitydata)
```
It is important to notice that the *date* variable is stored as a *factor*, rather than a *date*. 
Let's change it to a more easy to handle format
```{r}
activitydata$date <- as.Date(x = activitydata$date, format = "%Y-%m-%d")
str(activitydata)
```
In order to check the consistency, let's compute the number of *NA's* in the *steps* variable
```{r}
sum(is.na(activitydata$steps))
```
And the fraction of *NA's* is
```{r}
sum(is.na(activitydata$steps))/nrow(activitydata)
```

## What is mean total number of steps taken per day?
Use the library *dplyr* for data manipulation

```{r message = FALSE}
library(dplyr)

```
And then, the variable *totalsteps* is defined as follows, removing the *NA's* in the dataset using **na.rm=TRUE**  
```{r}
totalsteps <- activitydata %>% group_by(date) %>%  summarize(total = sum(steps, na.rm = TRUE))
head(totalsteps)
```
By using the *ggplot2* package, it is easier to create a plot using the grammar of graphics
```{r message = FALSE}
library(ggplot2)
```
And the Histogram for Total Steps per day is the following
```{r}
totalsteps %>% ggplot(aes(x = total)) + 
  geom_histogram(color = "white" ,fill = "orange",bins = 25) + 
  geom_vline(xintercept = mean(totalsteps$total, na.rm=TRUE), color = "blue", lwd = 1, lty = 2) +
  geom_vline(xintercept = median(totalsteps$total, na.rm = TRUE), color = "purple", lwd = 1, lty=3) +
  ggtitle("Histogram of total steps per day") +
  ylab("Count") + 
  xlab("Total steps per day") +
  theme_bw()
```  

Where the dashed line is the mean and the dotted line the median, with values:
```{r}
mean(totalsteps$total, na.rm = TRUE)

median(totalsteps$total, na.rm = TRUE)
```
## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
